Diversification of Australian blind snakes

### 2022-12-20 manuscript.Rmd
update manuscript

Based on 50 BioGeoBEARS biogeographic stochastic mappings (BSM) of the DEC+_j_+_x_ model. We estimated that an average of 71.2 ± 1.3 (mean ± SD) total anagenetic and cladogenetic events occurred throughout the phylogeny (see supplementary_table_BioGEOBEARS). Of these 40 ± 0 were cladogenetic events compared to 31.2 ± 1.3 anagenetic dispersal events. Out of the 40 cladogenetic dispersal events, most were narrow sympatry range expansion events (23.8±2.1 events) followed by subset sympatry (5.2±2.3 events), vicariance (4.9 ± 1.3), and founder events (6.1 ± 1.5).

@@

Based on current distributions (ALA) most _Anilios_ species are found entirely or at least partially in the tropical grassland biome (n = 20) followed by the arid biome (n = 13) and mediterranean biome (n = 10). The model indicates two pulses of biogeographic transitions for each  of the two major clades, which we will refer to as *Group 1* (n=21 lineages) and *Group 2* (n=14 lineages) (\@ref(fig:fig-biogeobears)a. The common ancestor of *Group 1* remained mostly in the Tropical Grassland Biome, then 12/21 lineages expanded or transitioned into other biomes. The common ancestor of *Group 2* first transitioned into the arid biome, and then 7/14 lineages expanded to the Mediterranean and Temperate Forest biomes, and/or transitioned back into the Tropical Grassland Biome.

@@

Consistent with @marinTracing2013, we found the initial diversification occurred very rapidly (<1 million years) in the tropical grassland biome followed by transitions and/or expansion into other biomes. 
However, our results show that species in both groups 1 and 2 have deep divergence between 20 – 10 MYA and gradually diversified during the time when the Australia was still relatively wet and humid [@byrneBirth2008] as opposed more recent invasion (<10 MYA) and diversification (<5 MYA) into the arid zone proposed by @marinHidden2013. This suggests that these lineages have diversified before the intensive continent-wide aridification in the late Miocene that resulted in the present dry and sandy areas of the arid biome [@byrneBirth2008].

One hypothesis to explain contemporary distribution in the arid biome is that *Anilios* may be pre-adapted to arid environments. Blind snakes have been found in xeric areas around the world, for example _Myriopholis macrorhyncha_ in the desert

North American  and the Palearctic _Letheobia episcopus_) possibly because subterranean microhabitat has relatively stable climate compared to surface [@kinlawReview1999;@nevoAdaptive1979].

We found that closely related species share similar biome and distribution. 
Our results indicate that cladogenetic processes (sympatry, vicariance, and founder event) made up the majority of range expansion events. Negative values for the distance parameter _X_ suggest that range expansion is expected to decrease as distance increase (CITE BioGeoBEARs intepetration?).

The fact that we observed multiple independent transitions into the arid biome.

For Group 1 eastward transitions might have occured

Pilbara in the west acted as a refugia and perhaps source of diversification 
Arid environemnts is heterogenous and some areas have acted as refugia during glaciation (CITE), which may cause isolation and subsequent speciation. For example, the Pilbara region has both lineages that have deep coalescent histories and shallower divergences (_A. ammodytes_ and _A. ganei_). These species may have once been widely distributed but became extirpated as aridification intensified and are now found only in their place of refuge during glacial fluctuations.

While there's not study to show, blind snakes probably have small ranges and poor dispersal ability as closely related species occupy the same or nearby areas. Stochastic events like floods may have facilitated dispersals but in short distances.

Australian blind snakes and these blind snakes share similar morphological features and diet consisting of eggs, larvae, and pupae. of ants and termites.

@@

Climate is one of the
- Australian blind snakes follow a different trajectory, one in which increased aridity from the late Miocene to early Pleistocene coincided with reduced diversification rates for genus. 
- In contrast to other vertebrates, our results suggest aridification reduces the rate of diversification in blind snakes. 
[@dornStrange2014;@deepakAridification2018;@catulloAridification2014;@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014;@owenHow2017;@crispHow2013;@toussaintUnveiling2015]
- Our study is the first to quantitatively assess the influence of aridity on the diversification rates of a group of vertebrate on a macroevolutionary scale.

- Overall the average temperature from Australia specific reconstruction is consistently is warmer than global mean temperature.

In the context of our birth-death models, our the absolute value of temperature might not have made much of a difference because our environment-dependent models fit the spline rather than absolute temperature.

- Paleoclimate reconstructions based on DEM by Scotese and Straume were similar in profile. However, DEM base on Valdes differed from the other two. Scotese is characterised by repeated fluctuations in the Pleistocene to reflect glacial cycles. 
However, models that take in account thermal niche tolerance, such as NEMo, will benefit from paleoclimate reconstructions such as these to inform limits (boundaries). 
- Aridity index is a good indicator, but aridity is geographically restricted to the arid zone. In our phylogeny, a lot of diversity is observed is in the mesic zones.
- Warmer and dryer environment may not directly increase speciation rate because blind snakes are physically adapting to warmer temperatures, rather it is the fragmentation of the landscape. 
- Thermal tolerance and preference for blind snakes remains unknown but fossorial species tend to live in pretty stable climate? 
- Our RPANDA model did not model environment-dependent extinction. 
- Studies that aim to understand how aridification affect diversification rates in Australia should apply paleoclimate or use them to inform models and inferences. 
- For example, models like NEMo allows ancestral niche to be informed by paleo environment as a prior.

## Rates of diversification between arid and non-arid lineages
- The early burst of speciation is observed in lineages that colonise new areas, ecological release. For example.

@@

- Interestingly blind snakes have diversified greatly compared to other fossorial vertebrates.

  - It is also possible that fossorial species are more cryptic and thus the full extent of their diversity is under appreciated. 
Microhabitat better explains diversification than climate, but typhlopidae has the second highest speciation rates among fossorial snakes [@bars-closelDiversification2017]. 
  - Extinction is a big cavaet especially because we suspect there are more species. Arid adapted lineages may have experienced higher extinction OR not well sampled in remote Western Australia. Most species are sampled next to roads (Xia's paper or do supplemental analyses).

  - Blind snakes are thought to not move very much and so formation of geological barriers during aridification may accelerate speciation between 20 – 10 MYA. 
  - High rates of diversification observed in concert with changing climate may be explained by allopatric speciation. 
  - Speciation could even be sympatric but no evidence to support this.

Future phylogeographic study should consider densely sampling species to see whether there are more expansion into or out of arid zone for widespread taxa. Two pulses of speciation early Miocene and then another early Pleistocene. Geographic barrier for mesic adapted lineages in the temperate and tropical forests may have a hard time crossing the lines [@bryantLines2016]

@@

@@

@@

- Limitations is we can only coarsely assign species to one biome one some are range-limited endemics on islands. For example *A. longissimus* has only been collected from 30m underground on Barrow Island located XX km the northwest coast of Western Australia. Placing that species with arid biome is not realistic but that area is considered the arid zone by our criteria.

@@

As we continue generate more genetic data to build ever more extensive phylogenies, there is a growing need to develop methods and complimentary paleo environmental datasets that allow us to test hypotheses about how different macroevolutionary processes influence diversification. We show that tools like `gen3sis` can be used to reconstruct localised paleo environment that provides a more appropriate fit to our Australian blind snake phylogeny than generic paleo global temperature. We believe these regional paleoclimate models will be of interested for studies where species are geographically restricted. 
- Furthermore we provide the first quantitative measure of paleo aridity

>>>>

Based on 50 BioGeoBEARS biogeographic stochastic mappings (BSM) of the DEC+_j_+_x_ model. We estimated that an average of 71.2 ± 1.3 (mean ± SD) total anagenetic and cladogenetic events occurred throughout the phylogeny (see supplementary_table_BioGEOBEARS). Of these 40 ± 0 were cladogenetic events compared to 31.2 ± 1.3 anagenetic dispersal events. Out of the 40 cladogenetic dispersal events, most were narrow sympatry range expansion events (23.8±2.1 events) followed by subset sympatry (5.2±2.3 events), vicariance (4.9 ± 1.3), and founder events (6.1 ± 1.5). Our model also indicate that range expansion is expected to decrease as distance increase (free parameter +_X_ = -1.3).

@@

Based on current distributions (ALA) most _Anilios_ species are found entirely or at least partially in the tropical grassland biome (n = 20) followed by the arid biome (n = 13) and mediterranean biome (n = 10). The model indicates two pulses of biogeographic transitions for each  of the two major clades, which we will refer to as *Group 1* (n=21 lineages) and *Group 2* (n=14 lineages) (\@ref(fig:fig-biogeobears)a. The common ancestor of *Group 1* remained mostly in the Tropical Grassland Biome, then 12/21 lineages expanded or transitioned into other biomes. The common ancestor of *Group 2* first transitioned into the arid biome, and then 7/14 lineages expanded to the Mediterranean and Temperate Forest biomes, and/or transitioned back into the arid biome.

@@

Consistent with previous research by @marinTracing2013, we found that the initial diversification of Anilios occurred in the Tropical Grassland biome, followed by transitions and/or expansion into other biomes. However, our analyses indicated that species-level divergence occurred earlier, between 20 – 10 MYA, compared to the 15 – 10 MYA proposed by @marinTracing2013.

Our BioGeoBEARS analyses suggest that _Anilios_ gradually diversified in the Tropical Grassland (Group 1) and arid biome (Group 2) during a time when Australia was relatively wet and humid [@byrneBirth2008]. This indicates that many species found in arid biomes, particularly in Group 2, diversified in the arid biome prior to the onset of aridification in the late Miocene, which has resulted in the dry and sandy landscape present today [@byrneBirth2008].

A hypothesis to explain the contemporary distribution of _Anilios_ and other blind snakes in arid biomes is that they may be pre-adapted to living in such environments. Blind snakes have been found in xeric areas around the world, such as _Rena humilis_ in the desert of North America and _Myriopholis macrorhyncha_ in semi-arid areas from east Africa to the Middle East [@sindacoReptiles2013]. To our knowledge, these species share similar dietary and morphological specialisations with their closest relatives, which is a pattern identified across all fossorial snakes [@cyriacDigging2018]. Because the subterranean microhabitat has a relatively stable climate compared to the surface [@kinlawReview1999;@nevoAdaptive1979], modification from the already specialised phenotype may not be selected for. Future studies that investigate phenotypic variations across all blind snakes will further our understanding of how these snakes evolved with seemingly conserved phenotypes.

@@

Changes in paleoenvironmental changes have been linked with diversification rates via range contraction and expansion that influences niche availability. 
We found that aridity played an important role in 
- Australian blind snakes follow a different trajectory, one in which increased aridity from the late Miocene to early Pleistocene coincided with reduced diversification rates for genus. This is in contrast with other vertebrates, where diversification rates seem to increase with aridification [@dornStrange2014;@deepakAridification2018;@catulloAridification2014;@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014;@owenHow2017;@crispHow2013;@toussaintUnveiling2015]
- Warmer and dryer environment may not directly increase speciation rate because blind snakes are physically adapting to warmer temperatures. However, climatic fluctuations can contract and fragment available niche for blind snakes andthus influence diversification rates. 
- In the context of our birth-death models, the absolute value of temperature might not have made much of a difference because our environment-dependent models fit the spline rather than absolute temperature. However, models that take in account thermal niche tolerance, such as NEMo [CITE], will benefit from paleoclimate reconstructions such as these to inform limits (boundaries). 
We suggest future studies that aim to understand how aridification affect diversification rates in Australia consider using these paleoclimate models and/or use them to inform inferences.

ALEX SKEELS add to this discussion?
- Paleoclimate reconstructions based on DEM by Scotese and Straume were similar in profile. However, DEM base on Valdes differed from the other two. Scotese is characterised by repeated fluctuations in the Pleistocene to reflect glacial cycles. to add discussion to this?

## Rates of diversification between arid and non-arid lineages

- The early burst of speciation is observed in lineages that colonise new areas. For example.

@@

- Interestingly blind snakes have diversified greatly compared to other fossorial vertebrates. 
  - It is also possible that fossorial species are more cryptic and thus the full extent of their diversity is under appreciated.

Microhabitat better explains diversification than climate, but typhlopidae has the second highest speciation rates among fossorial snakes [@bars-closelDiversification2017]. 
  - Extinction is a big cavaet especially because we suspect there are more species. Arid adapted lineages may have experienced higher extinction OR not well sampled in remote Western Australia. Most species are sampled next to roads [CITE Xia's paper or do supplemental analyses].

- Vast arid biome is  the largest biome in Australia with lots of heterogeneity (cite Mitzy's paper). Pockets of certain areas were refugia during glaciation cycles of the Pleistocene.

- Future phylogeographic study should consider densely sampling species to see whether there are more expansion into or out of arid zone for widespread taxa. 
- Geographic barrier for mesic adapted lineages in the temperate and tropical forests may have difficulty crossing the lines [@bryantLines2016]

@@

- Limitations is we can only coarsely assign species to one biome one some are range-limited endemics on islands. For example *A. longissimus* has only been collected from 30m underground on Barrow Island located XX km the northwest coast of Western Australia. Placing that species with arid biome is not realistic but that area is considered the arid biome by the WWF Ecoregions map.

@@

- Our RPANDA model did not model environment-dependent extinction.

@@

@@

As we continue generate more genetic data to build ever more extensive phylogenies, there is a growing need to develop methods and complimentary paleo environmental datasets that allow us to test hypotheses about how different macroevolutionary processes influence diversification. We show that tools like `gen3sis` can be used to reconstruct localised paleoaridity data that provides a better fit to our Australian blind snake phylogeny than generic global paleotemperature. We believe these regional paleoclimate models will be of interested for studies where species are geographically restricted.
